### 2015-06-19 doc/thesis/system.tex
Back porting table refinement from paper to thesis

Back to first-level types, table types represent the various forms
that Lua tables can take.
The syntactical form of table types is $\{\overline{F{:}V}\}_{unique|open|fixed|closed}$,
where the notation $\overline{F{:}V}$ denotes the list $F_{1}{:}V_{1}, ..., F_{n}{:}V_{n}$.
Each $V_{i}$ represents the type of the value that table keys of type $F_{i}$ map to.
%Key types can only be literal types, base types, or the top type.
%We made this restriction to the type of the keys because the statistics
%that we discussed in Section \ref{sec:statistics} showed that most
%of the tables are records, lists, and hashes.
%The type $\Value$ is an option when we need a loose table type.
%For instance, $\{\Value:\Value\}_{closed}$ represents the type of a
%table in which both indices and values can have any type.
Value types represent mutable values by default, but we can use the
$\Const$ type to make them represent immutable values.

@@

The simplest expression involving tables is the empty table constructor.
Its type checking rule is straightforward:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR1}\\
\env_{1}, \penv \vdash \{\}:\{\}_{unique}, \env_{1}
\end{array}
\]

@@

First, our type system uses the auxiliary relation
$\env_{1}, \penv \vdash [e_{1}] = e_{2} : (K,V), \env_{2}$ to type check each
table field.
This auxiliary relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, checking a table field $[e_{1}] = e_{2}$
produces a pair $(K,V)$ and a new type environment $\env_{2}$.
A pair $(K,V)$ means that $e_{1}$ has type $K$ and $e_{2}$ has type $V$,
where $K$ is the type of the key and $V$ is the type of the value.

After type checking each table field, our type system uses each pair $(K,V)$
to build the table type that express the type of a given constructor, and
uses the predicate \emph{wf} to check whether this table type is well-formed.
Formally, a table type is well-formed if it obeys the following rule:
\[
\forall i \not\exists j \; i \not= j \wedge K_{i} \lesssim K_{j}
\]

@@

Well-formed table types also do not allow \emph{unique} and
\emph{open} table types to appear in the type of the values.
We made this restriction because our type system does not keep
track of aliases to table fields.
This means that allowing \emph{unique} and \emph{open} table
types to appear in the type of a value would allow the
creation of unsafe aliases.
The rules that type check table fields use the auxiliary function
\emph{close} to close the type of the values in a table type.

The rule \textsc{T-CONSTRUCTOR2} uses these steps to type check a
table constructor with a non-empty list of table fields:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(K_{i},V_{i}), \env_{i+1} \\
       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
       wf(T) \;\;\;
       n = |\;\overline{[e_{1}] = e_{2}}\;| \\
       \env_{f} = merge(\env_{1}, ..., \env_{n+1})
       \end{array}}
      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{f}}
\end{array}
\]

@@

As another example, the constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$ through rule
\textsc{T-CONSTRUCTOR2}.
The inner table is \emph{closed} to prevent the creation of unsafe aliases.

@@

In this example, we assign the type $\{\}_{unique}$ to the variable
$person$, then we refine its type to $\{``firstname":\String\}_{unique}$,
and then we refine its type to $\{``firstname":\String, ``lastname":\String\}_{unique}$.
Rule \textsc{T-REFINE} type checks this use of refinement:
\[
\begin{array}{c}
\mylabel{T-REFINE}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n} \}_{open|unique}\\
       \env_{1}, \penv \vdash k:K, \env_{2} \;\;\;
       \not \exists i \in 1..n \; K \lesssim K_{i} \;\;\;
       V = close(T)
       \end{array}}
      {\env_{1}, \penv \vdash id[k] {<}T{>}:V, \env_{2}[id \mapsto \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}, K{:}V\}_{open|unique}]}
\end{array}
\]

We restricted the refinement of table types to include only literal
keys, because its purpose is to make it easier the construction of
table types that represent records.

@@

We also do not allow the refinement of table types to introduce
fields with table types that are not \emph{closed}.
For instance,
\begin{center}
\begin{tabular}{l}
$\string_ENV[``x"] \; {<}\{\}_{unique}{>} = \{\}$
\end{tabular}
\end{center}
refines the type of $\string_ENV$ from $\{\}_{open}$ to $\{``x":\{\}_{closed}\}_{open}$.
Currently, our type system can only track \emph{unique} and
\emph{open} table types that are bound to local variables.

@@

This example type checks because all the environment changes are consistent, and
$``foo" \times 1 \times \Nil{*} \lesssim \String \times \Integer \times \Value{*}$.
By consistent we mean that we are only adding new fields.
Nevertheless, the next example does not type check because it tries to add
the same field to $\string_ENV$, but with different types:
\[
\string_ENV[``x"] \; {<}\String{>}, \string_ENV[``x"] \; {<}\Integer{>} = ``foo", 1
\]

@@

Rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this non-deterministic behavior.
Rule \textsc{T-IDREAD1} uses the auxiliary function \emph{close} to
produce a \emph{closed} alias.
It also uses the auxiliary function \emph{open} to change the type of
the original reference from \emph{unique} to \emph{open},
because aliasing an \emph{unique} table type while keeping the original
reference \emph{unique} can be unsafe.
Rule \textsc{T-IDREAD2} uses the auxiliary function \emph{fix} to
produce a \emph{fixed} alias.
It also uses \emph{fix} to change the type of the original reference
to \emph{fixed}, because a \emph{fixed} table type does not allow
width subtyping.
We define these rules as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-IDREAD1}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env_{1}, \penv \vdash id:close(T_{2}), \env_{1}[id \mapsto open(T_{1})]}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-IDREAD2}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env_{1}, \penv \vdash id:fix(T_{2}), \env_{1}[id \mapsto fix(T_{1})]}
\end{array}
\end{array}
\]

Both rules use the auxiliary function \emph{read} because they may be
accessing an identifier that is bound to a filter or projection type.
As we mentioned in Section \ref{sec:unions}, our type system includes
a small set of \texttt{type} predicates that allow programmers to
discriminate union types, and our type system uses filter and projection
types in the definition of these predicates to handle the discrimination
of unions types.
While filter types discriminate unions of first-level types, projection
types discriminate unions of second-level types and project unions of
second-level types into unions of first-level types.
We can define \emph{read} as follows:
\begin{align*}
read(\penv, \phi(T_{1},T_{2})) & = T_{2}\\
read(\penv, \pi_{i}^{x}) & = proj(\penv(x), i)\\
read(\penv, T) & = T
\end{align*}

The function \emph{read} uses the auxiliary function \emph{proj}
to project a union of first-level types, based on an union of
second-level types and an index from a projection type.
In Section \ref{sec:fap} we will discuss how our type system uses
projection types to handle overloaded return types.
We can define \emph{proj} as follows:
\begin{align*}
proj(T_{1} \times ... \times T_{n} \times T{*}, i) & =  T_{i} \;\;\; \text{if $i <= n$}\\
proj(S_{1} \sqcup S_{2}, i) & = proj(S_{1}, i) \cup proj(S_{2}, i)
\end{align*}

We also need to close \emph{unique} and \emph{open} tables that
appear in the left-hand side of assignments, as leaving them
\emph{unique} and \emph{open} would allow the creation of
unsafe references.

As an example,
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = \{\} \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$b = a;$}\\
& & \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& & \multicolumn{1}{l}{$b[``x"] \; {<}\Integer{>} = 1$}\\
\end{tabular}
\end{center}
does not type check because we cannot add the field $``x"$ to $b$,
as its type is \emph{closed}.
Aliasing $a$ changes the type of $a$ from $\emph{unique}$ to
$\emph{open}$, and that is the reason why we can add the field
$``x"$ to the type of $a$.
Aliasing $a$ also produces the type $\{\}_{closed}$, which is
the same type that $b$ has in left-hand side of the assignment.
After the assignment, the type of $b$ is \emph{closed} and thus
does not allow changing the value that is stored in the field
$``x"$ of $a$.

Rule \textsc{T-IDWRITE} defines this behavior:
\[
\begin{array}{c}
\mylabel{T-IDWRITE}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = write(T_{1})}
      {\env_{1}, \penv \vdash id_{l}:close(T_{2}), \env_{1}[id \mapsto close(T_{2})]}
\end{array}
\]

This rule uses the auxiliary function \emph{write} because it may be
accessing an identifier that is bound to a filter type.
As we mentioned in Sections \ref{sec:unions} and \ref{sec:alias},
assignments restore discriminated union types to their original types,
and function \emph{write} works in this purpose.
We can define \emph{write} as follows:
\begin{align*}
write(\phi(T_{1},T_{2})) & = T_{1}\\
write(T) & = T
\end{align*}

Our type system also has different rules for type checking table indexing to avoid
changing table types in these operations, as they cannot create aliases:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-INDEX1}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = T \;\;\;
       read(\penv, T) = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}\\
       \env_{1}, \penv \vdash e_{2}:K, \env_{2} \;\;\;
       \exists i \in 1{..}n \; K \lesssim K_{i}
       \end{array}}
      {\env_{1}, \penv \vdash id[e_{2}]:V_{i}, \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-INDEX2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash e_{1}:\{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}, \env_{2}\\
       \env_{2}, \penv \vdash e_{2}:K, \env_{3} \;\;\;
       \exists i \in 1{..}n \; K \lesssim K_{i}
       \end{array}}
      {\env_{1}, \penv \vdash e_{1}[e_{2}]:V_{i}, \env_{3}}
\end{array}
\end{array}
\]

A second form of refinement happens when we want to use an
\emph{unique} or \emph{open} table type in a context that expects a
\emph{fixed} or \emph{closed} table type with a different shape.
This kind of refinement allows programmers to add optional fields
or merge existing fields.
To do that, Typed Lua includes a type coercion expression ${<}T{>} \; id$.
For instance, we can use this type coercion expression to make the following
example type check:
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{ \} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& \multicolumn{2}{l}{$a[``y"] \; {<}\String{>} = ``bar";$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``x":\String, ``y":\String \cup \Nil \}_{closed} =$}\\
& & \multicolumn{1}{l}{${<}\{``x":\String, ``y":\String \cup \Nil\}_{open}{>} \; a \; \mathbf{in} \; a[``z"] \; {<}\Integer{>} = 1$}
\end{tabular}
\end{center}

@@

Rule \textsc{T-COERCE} defines the behavior of the type coercion expression:
\[
\begin{array}{c}
\mylabel{T-COERCE}\\
\dfrac{\env_{1}(id) \subtype T \;\;\;
       \env_{1}[id \mapsto T], \penv \vdash id:T_{1}, \env_{2}}
      {\env_{1}, \penv \vdash {<}T{>} \; id:T_{1}, \env_{2}}
\end{array}
\]

Note that rule \textsc{T-COERCE} only allows changing the type
of a variable if the new type is a supertype of the previous type,
and the resulting type is always \emph{fixed} or \emph{closed}
to prevent the creation of unsafe aliases.

We also need to make sure to close all \emph{unique} and \emph{open}
table types before we type check a nested scope.
To do that, our type system uses some auxiliary functions to change
the type of variables before type checking a nested scope and
also to change the type of assigned and referenced variables after
type checking a nested scope.
The function \emph{crall} closes all \emph{unique} and \emph{open}
table types; it also restores filter types to their original types.
The function \emph{crset} closes a given set of free assigned variables,
which is given by the function \emph{fav}, and
it also uses this set to restore filter types to their original types.
The function \emph{openset} changes from \emph{unique} to \emph{open}
a given set of referenced variables, which is given by the function \emph{rv}.

As an example,
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique}, b:\{\}_{unique} = \{\}, \{\} \; \mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{local} \; f:\Integer \times \Void \rightarrow \Integer \times \Void =$}\\
& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer):\Integer \times \Void$}\\
& & & \multicolumn{1}{l}{$b = a \;;\; \mathbf{return} \; x + 1$}\\
& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;\; b[``x"] \; {<}\String{>} = ``foo" \;;\; f(a[``x"])$}
\end{tabular}
\end{center}
does not type check because we cannot add the field
$``x"$ to $b$, as its type is closed.
The assignment $b = a$ type checks because, at that point,
$a$ and $b$ have the same type: $\{\}_{closed}$.
Their type was closed by \emph{crall} before type checking
the function body.
Their type would be restored to $\{\}_{unique}$ after type checking
the function body, but that assignment also triggers other two type changes.
First, the function \emph{fav} includes $b$ in the set of variables
that should be closed by \emph{crset}.
Then, the function \emph{rv} includes $a$ in the set of variables
that should change from \emph{unique} to \emph{open} by \emph{openset}.
After declaring $f$, $a$ has type $\{\}_{open}$ and $b$ has type $\{\}_{closed}$,
so we can refine the type of $a$, but we cannot refine the type of $b$.

Rule \textsc{T-FUNCTION1} illustrates this case:
\[
\begin{array}{c}
\mylabel{T-FUNCTION1}\\
\dfrac{\begin{array}{c}
       crall(\env_{1}[\overline{id} \mapsto \overline{T}]), \penv[\ret \mapsto S] \vdash s, \env_{2}\\
       \env_{3} = openset(crset(\env_{1}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s)),rv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
       \end{array}}
      {\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}T}){:}S \; s:\overline{T} \times \Void \rightarrow S, \env_{3}}
\end{array}
\]

>>>>

Back to first-level types, table types represent the various forms
that Lua tables can take.
The syntactical form of table types is $\{\overline{F{:}V}\}_{unique|open|fixed|closed}$,
where the notation $\overline{F{:}V}$ denotes the list $F_{1}{:}V_{1}, ..., F_{n}{:}V_{n}$.
Each $V_{i}$ represents the type of the value that table keys of type $F_{i}$ map to.
Value types represent mutable values by default, but we can use the
$\Const$ type to make them represent immutable values.

@@

The simplest expression involving tables is the empty table constructor $\{\}$;
it always has type $\{\}_{unique}$.

@@

First, our type system uses the auxiliary relation
$\env_{1}, \penv \vdash [e_{1}] = e_{2} : (F,V), \env_{2}$ to type check each
table field.
This auxiliary relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, checking a table field $[e_{1}] = e_{2}$
produces a pair $(F,V)$ and a new type environment $\env_{2}$.
A pair $(F,V)$ means that $e_{1}$ has type $F$ and $e_{2}$ has type $V$,
where $F$ is the type of the key and $V$ is the type of the value.

After type checking each table field, our type system uses each pair $(F,V)$
to build the table type that express the type of a given constructor, and
uses the predicate \emph{wf} to check whether this table type is well-formed.
Formally, a table type is well-formed if it obeys the following rule:
\[
\begin{array}{rcl}
wf(\{\overline{F:V}\}_{unique|open|fixed|closed}) & = & \forall i \; ((\nexists j \; i \not= j \,\wedge\, F_{i} \lesssim F_{j}) \,\wedge\, wf(V_{i}) \,\wedge\\
& & \;\;\;\; \lnot tag(V_{i},unique) \,\wedge\, \lnot tag(V_{i},open))\\
wf({\bf const}\; F) & = & wf(F) \\
wf(F_1 \cup F_2) & = & wf(F_1) \,\wedge\, wf(F_2) \\
wf(\mu x.F) & = & wf(F) \\
wf(S_1 \rightarrow S_2) & = & wf(S_1) \,\wedge\, wf(S_2) \\
wf(S_1 \sqcup S_2) & = & wf(S_1) \,\wedge\, wf(S_2)\\
wf(F{*}) &=& wf(F) \\
wf(F \times P) &=& wf(F) \,\wedge\,wf(P)\\
wf(F) & = & \top \;\;\;\mathrm{for\; all\; other\; cases}\\\\
tag(F_1 \cup F_2, t) & = & tag(F_1, t) \,\vee\, tag(F_2,t) \\
tag(\{\overline{F:V}\}_t, t) & = & \top\\
\end{array}
\]

@@

Well-formed table types also do not allow \emph{unique} and
\emph{open} table types to appear in the type of the values.
We made this restriction because our type system does not keep
track of aliases to table fields.
This means that allowing \emph{unique} and \emph{open} table
types to appear in the type of a value would allow the
creation of unsafe aliases.
The rules that type check table fields use the auxiliary function
\emph{fix} to change any \emph{unique} and \emph{open} table types
used in the field initializer to \emph{fixed}.

The rule \textsc{T-CONSTRUCTOR2} uses these steps to type check a
table constructor with a non-empty list of table fields:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(F_{i},V_{i}), \env_{i+1} \\
       T = \{F_{1}{:}V_{1}, ..., F_{n}{:}V_{n}\}_{unique} \;\;\;
       wf(T) \;\;\;
       n = |\;\overline{[e_{1}] = e_{2}}\;| \\
       \env_{f} = merge(\env_{1}, ..., \env_{n+1})
       \end{array}}
      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{f}}
\end{array}
\]

@@

As another example, the constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
has type $\{``x":1, ``y":\{``z":2\}_{fixed}\}_{unique}$ through rule
\textsc{T-CONSTRUCTOR2}.
The inner table is \emph{fixed} to prevent the creation of unsafe aliases.

@@

In this example, we assign the type $\{\}_{unique}$ to the variable
$person$, then we refine its type to $\{``firstname":\String\}_{unique}$,
and then we refine its type to $\{``firstname":\String, ``lastname":\String\}_{unique}$.
Rule \textsc{T-REFINE1} type checks this use of refinement:
\[
\begin{array}{c}
\mylabel{T-REFINE1}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \{\overline{F{:}V}\}_{unique}\\
       \env_{1}, \penv \vdash e:F_{new}, \env_{2} \;\;\;
       \nexists i \in 1..n \; F_{new} \lesssim F_{i} \;\;\;
       V_{new} = vt(F_{new},V) \;\;\; n = |\overline{F{:}V}|
       \end{array}}
      {\env_{1}, \penv \vdash id[e] {<}V{>}:V_{new}, \env_{2}[id \mapsto \{\overline{F{:}V}, F_{new}{:}V_{new}\}_{unique}]}
\end{array}
\]

The rule \textsc{T-REFINE} uses the auxiliary function \emph{vt}
to adjust the type of the value according to the type of the key.
More precisely, \emph{vt} includes the type $\Nil$ in the type of
the value when the type of the key is not a literal type.
It is also this function that uses \emph{fix} to prevent
\emph{unique} and \emph{open} table types to appear in the type
of the values.

@@

We also do not allow the refinement of table types to introduce
fields with table types that are neither \emph{fixed} nor \emph{closed}.
For instance,
\begin{center}
\begin{tabular}{l}
$\string_ENV[``x"] \; {<}\{\}_{unique}{>} = \{\}$
\end{tabular}
\end{center}
refines the type of $\string_ENV$ from $\{\}_{open}$ to $\{``x":\{\}_{fixed}\}_{open}$.
Currently, our type system can only track \emph{unique} and
\emph{open} table types that are bound to local variables.

@@

This example type checks because all the environment changes are consistent, and
$``foo" \times 1 \times \Nil{*} \lesssim \String \times \Integer \times \Value{*}$.
By consistent we mean that we are only adding new fields.
More precisely, the first coercion expression refines the type of $\string_ENV$
to $\{``x":\String\}_{open}$, while
the second coercion expression refines the type of $\string_ENV$
to $\{``y":\Integer\}_{open}$.
Merging the two yields $\{``x":\String, ``y":\Integer\}_{open}$.
Nevertheless, the next example does not type check because it tries to add
the same field to $\string_ENV$, but with different types:
\[
\string_ENV[``x"] \; {<}\String{>}, \string_ENV[``x"] \; {<}\Integer{>} = ``foo", 1
\]

@@

Rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this non-deterministic behavior.
Rule \textsc{T-IDREAD1} uses the auxiliary function \emph{close} to
produce a \emph{closed} alias.
It also uses the auxiliary function \emph{open} to change the type of
the original reference from \emph{unique} to \emph{open},
because aliasing an \emph{unique} table type while keeping the original
reference \emph{unique} can be unsafe.
Rule \textsc{T-IDREAD2} uses the auxiliary function \emph{fix} to
produce a \emph{fixed} alias.
It also uses \emph{fix} to change the type of the original reference
to \emph{fixed}, because a \emph{fixed} table type does not allow
width subtyping.
We define these rules as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-IDREAD1}\\
\dfrac{\env_{1}(id) = F}
      {\env_{1}, \penv \vdash id:close(F), \env_{1}[id \mapsto open(F)]}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-IDREAD2}\\
\dfrac{\env_{1}(id) = F}
      {\env_{1}, \penv \vdash id:fix(F), \env_{1}[id \mapsto fix(F)]}
\end{array}
\end{array}
\]

We do not need to close \emph{unique} and \emph{open} tables that
appear in the left-hand side of assignments, because \textsc{T-IDREAD}
is sufficient to forbid the creation of an alias to another \emph{unique}
or \emph{open} table.
For this reason, identifiers that appear in the left-hand side
of assignements have their own rule \textsc{T-IDWRITE}:
\[
\begin{array}{c}
\mylabel{T-IDWRITE}\\
\dfrac{\env_{1}(id) = F}
      {\env_{1}, \penv \vdash id_{l}:F, \env_{1}}
\end{array}
\]

Our type system also has different rules for type checking table indexing to avoid
changing table types in these operations, as they cannot create aliases.
These rules also use the auxiliary function \emph{rconst} to strip the
$\Const$ type from the type of the value, if present.
We define these rules as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-INDEXREAD1}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \{\overline{F{:}V}\} \;\;\;
       \env_{1}, \penv \vdash e_{2}:F, \env_{2} \;\;\;
       \exists i \in 1{..}n \; F \lesssim F_{i} \;\;\;
       n = |\overline{F{:}V}|
       \end{array}}
      {\env_{1}, \penv \vdash id[e_{2}]:rconst(V_{i}), \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-INDEXREAD2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash e_{1}:\{\overline{F{:}V}\}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{2}:F, \env_{3} \;\;\;
       \exists i \in 1{..}n \; F \lesssim F_{i} \;\;\;
       n = |\overline{F{:}V}|
       \end{array}}
      {\env_{1}, \penv \vdash e_{1}[e_{2}]:rconst(V_{i}), \env_{3}}
\end{array}
\end{array}
\]

Rule \textsc{T-INDEXREAD1} defines the case where using an identifier
to index a table does not create an alias, while rule \textsc{T-INDEXREAD2}
defines the case for indexing expressions where the expression
denoting the table is not an identifier.
The rules for indexing left-hand values are similar to these rules,
except that they ensure that the field is not $\Const$.

A second form of refinement happens when we want to use an
\emph{unique} or \emph{open} table type in a context that expects a
\emph{fixed} or \emph{closed} table type with a different shape.
This kind of refinement allows programmers to add optional fields
or merge existing fields.
To do that, Typed Lua includes a type coercion expression ${<}F{>} \; id$.
For instance, we can use this type coercion expression to make the following
example type check:
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{ \} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& \multicolumn{2}{l}{$a[``y"] \; {<}\String{>} = ``bar";$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``x":\String, ``y":\String \cup \Nil \}_{closed} =$}\\
& & \multicolumn{1}{l}{${<}\{``x":\String, ``y":\String \cup \Nil\}_{open}{>} \; a \; \mathbf{in} \; a[``z"] \; {<}\Integer{>} = 1$}
\end{tabular}
\end{center}

@@

Rules \textsc{T-COERCE1} and \textsc{T-COERCE2} define the behavior of the
type coercion expression:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-COERCE1}\\
\dfrac{\env_{1}(id) \subtype F \;\;\; tag(F,closed)}
      {\env_{1}, \penv \vdash {<}F{>} \; id:F, \env_{1}[id \mapsto reopen(F)]}
\end{array}
\;
\begin{array}{c}
\mylabel{T-COERCE2}\\
\dfrac{\env_{1}(id) \subtype F \;\;\; tag(F,fixed)}
      {\env_{1}, \penv \vdash {<}F{>} \; id:F, \env_{1}[id \mapsto F]}
\end{array}
\end{array}
\]

Note that the coercion rules only allows changing the type
of a variable if the new type is a supertype of the previous type,
and the resulting type is always \emph{fixed} or \emph{closed}
to prevent the creation of unsafe aliases.

We also need to make sure to close all \emph{unique} and \emph{open}
table types before we type check a nested scope.
To do that, our type system uses some auxiliary functions to change
the type of variables before type checking a nested scope and
also to change the type of assigned and referenced variables after
type checking a nested scope.
The function \emph{closeall} closes all \emph{unique} and \emph{open} table types.
The function \emph{closeset} closes a given set of free assigned variables,
which is given by the function \emph{fav}.
The function \emph{openset} changes from \emph{unique} to \emph{open}
a given set of referenced variables, which is given by the function \emph{frv}.

As an example,
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique}, b:\{\}_{unique} = \{\}, \{\} \; \mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{local} \; f:\Integer \times \Nil{*} \rightarrow \Integer \times \Nil{*} =$}\\
& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer):\Integer \times \Nil{*}$}\\
& & & \multicolumn{1}{l}{$b = a \;;\; \mathbf{return} \; x + 1$}\\
& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;\; b[``x"] \; {<}\String{>} = ``foo" \;;\; f(a[``x"])$}
\end{tabular}
\end{center}
does not type check because we cannot add the field
$``x"$ to $b$, as its type is closed.
The assignment $b = a$ type checks because, at that point,
$a$ and $b$ have the same type: $\{\}_{closed}$.
Their type was closed by \emph{closeall} before type checking
the function body.
Their type would be restored to $\{\}_{unique}$ after type checking
the function body, but that assignment also triggers other two type changes.
First, the function \emph{fav} includes $b$ in the set of variables
that should be closed by \emph{closeset}.
Then, the function \emph{frv} includes $a$ in the set of variables
that should change from \emph{unique} to \emph{open} by \emph{openset}.
After declaring $f$, $a$ has type $\{\}_{open}$ and $b$ has type $\{\}_{closed}$,
so we can refine the type of $a$, but we cannot refine the type of $b$.

Rule \textsc{T-FUNCTION1} illustrates this case:
\[
\begin{array}{c}
\mylabel{T-FUNCTION1}\\
\dfrac{\begin{array}{c}
       closeall(\env_{1})[\overline{id \mapsto F}], \penv[\ret \mapsto S] \vdash s, \env_{2} \\
       \env_{3} = openset(\env_{1}, frv(\mathbf{fun} \; (\overline{id{:}F}){:}S \; s)) \\
       \env_{4} = closeset(\env_{3}, fav(\mathbf{fun} \; (\overline{id{:}F}){:}S \; s))
       \end{array}}
      {\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}F}){:}S \; s:F_{1} \times ... \times F_{n} \times \Nil{*} \rightarrow S, \env_{4}}
\end{array}
\]
